Addressing climate change and development pressures in an urban estuary through habitat restoration planning

### 2022-09-08 manu-draft.qmd
proofing intro, methods, updating workflow graphic

Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise associated with climate change. To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of the habitat mosaic, and anticipated future impacts from coastal stressors. A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short-term (2030) targets and long-term (2050) goals. The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region. Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions. The new paradigm also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida. The resulting habitat goals are spatially explicit with maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands. This approach represents a general framework to support coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments. Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.

@@

The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration or conservation activities in coastal environments. Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed. Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09]. The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting the variety of ecosystem services provided by these habitats and the wildlife they support [@Enwright15]. Given projected habitat losses without management intervention and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.

Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs among public and private sectors, nor do they fully account for anticipated effects of multiple stressors. For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems by addressing issues within hydrologically-defined geographic areas. Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects has been recognized as an effective means of allowing ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09]. Adaptive management [@Holling78; @Gregory06] components have been increasingly used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans. Elements of each of these approaches could be combined to create a more holistic approach to guide restoration and conservation activities for coastal habitats in urban settings.

The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration and conservation plan that addresses pervasive coastal stressors. Compared to other estuaries, the ratio of watershed to estuary area is small and the area is heavily developed with 42% of land use classified as urban and suburban residential [@SWFWMD18]. A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously employed [@Lewis96; @Robison10; @Cicchetti11; @Russell15]. Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period. Criticisms of this approach included a lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts. These challenges are shared by restoration practitioners in other coastal environments and an approach that accommodates these challenges for planning would be highly transferable.

This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Relatively consistent data are available for most Tampa Bay habitats of interest (1988 to 2018), representing a time period when federal, state and local regulations were in effect and regional impacts from climate change are documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the current (2020) ecotone for mangrove and salt marsh habitats. The upstream watershed includes multiple habitats, including pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties of the watershed [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

The major habitat types of Tampa Bay were stratified relative to tidal influence and location in the watershed. Subtidal habitats included those that are submerged all or most of the time; intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides; and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).

Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habiats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @JEI16; @JEI20; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour area was considered the most likely to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase from 58 to 259 cm in the region [@Burke19].

@@

Opportunity areas for restoration and recommended coverage targets for habitat types were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the Tampa Bay watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats. Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 source data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19]. Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) geospatial database [@SWFWMD18]. Land use and cover types (natural and developed) are classified pursuant to the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data. Therefore, all applicable FLUCCS codes representing the suite of natural freshwater wetlands were combined within those two classifications. Native upland habitats were also combined in one classification. Historical estimates for these habitat coverages were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.

To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates. These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines, only available as current estimates with no information on historical trends.

Current and historical coverage estimates were obtained for all habitat types, including estimates of developed lands, for each year of available data spanning approximately three decades at two-year time steps. In addition, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types. This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats). The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].

@@

Past restoration projects were quantified for each of the major habitat types, when available, to inform expectations for setting short-term goals and long-term targets. Information regarding habitat restoration and enhancement activities in the Tampa Bay area over the past 40 years was compiled from various sources. Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock). Enhancement was defined as any activity not including earthwork that improved the environment (e.g., planting native vegetation, invasive species or debris removal, prescribed burns, etc.). Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, size (area or length), and location (latitude and longitude). Data gaps were supplemented by archival research, site visits, contacting the responsible entities, and expert knowledge form local professionals. Living shoreline projects, including seawall enhancements and oyster reef installations, were inventoried separately.

@@

Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow. The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify restoration opportunities. Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain defined goals and targets. The definition and mapping of opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible under current and future projected conditions, related to anticipated land use change and sea-level rise.

The Land Use Land Cover 2017 dataset from SWFWMD was used for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata. All FLUCCS classification codes were placed into one of three categories. First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed, and were further grouped into three major habitat types (tidal wetlands, freshwater wetlands, and uplands). Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration. Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.

After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas. Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed. The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum. This area extends from the local Mean Lower Low Water (MLLW) elevation to elevation 5 feet above Mean Sea Level and is likely to be affected by frequent tidal flooding or inundation [@Burke19]. Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.

The existing and proposed native and restorable areas were those that occured in public lands and parcels currently acquired or targeted for acquisition. To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory, consulting staff from various federal, state and local entities, and inventorying conservation and drainage easements data. This workflow created the existing conservation and proposed (i.e., planned to be acquired) conservation layers in @fig-gisworkflow . Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.

@@

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential. @Ries14 created a soils suitability analysis for wetland mitigation and restoration and classified all soils in the Tampa Bay watershed as xeric, mesic, or hydric. The mesic and hydric categories were combined to represent wetland restoration potential, while the xeric category was used to represent upland restoration potential. A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, while mesic or hydric soils occurring above the 5-foot contour were assigned a restoration potential for freshwater wetlands.

Two distinct mapping products were created from the above analysis. The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable. The second was a map that identified the restorable lands (either proposed, existing, or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands (which included salt marshes). All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18]. The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) for use by others [@Beck22].

>>>>

Native habitats in Florida face dual pressures at the land-sea interface from urban development and sea-level rise. To address these pressures, restoration practitioners require robust tools that identify reasonable goals given historical land use trends, current status of native habitats, and anticipated future impacts from coastal stressors. A target-setting approach for native habitats was created for the Tampa Bay watershed that identifies current restoration opportunities and establishes short-term (2030) targets and long-term (2050) goals. The approach was informed through a three-decade habitat change analysis and over forty years of habitat restoration projects in the region. Restoration goals were defined based on what is possible today and the projected needs for the future, rather than attempting to replicate past ecological conditions. The new approach also accounts for the expected impacts of sea-level rise, climate change, and watershed development - stressors which are pervasive in Florida. The resulting habitat goals are spatially explicit with maps that identify remaining restoration opportunities, while also providing an approach for the entire watershed that targets subtidal, intertidal, and uplands. This approach also supports coastal planning decisions that need to address competing interests and could be applied in other coastal settings where sustainable urbanization practices need to co-exist with natural environments. Methods for repeatable analyses are also available using an open source workflow to update progress over time and for adoption by others.

@@

The combined effects of land development and climate change are especially problematic for prioritizing habitat restoration activities in coastal environments. Landward migration of critical habitats in response to sea-level rise may not be possible due to anthropogenic barriers in the watershed. Sea-level rise can occur quicker than landward migration of salt marshes and the upland slope may already be lost to urban development and hardening [@Titus09]. The coastal squeeze of development and climate change diminishes the available space for future restoration in urban estuaries, thereby impacting ecosystem services provided by these habitats and the wildlife they support [@Enwright15]. Given projected habitat losses and the limited resources available, appropriate and realistic sites for restoration need to be identified that account for future stressors and past trends.

Past approaches for guiding restoration planning have been successfully used in other contexts, but they do not fully balance competing needs among public and private sectors, nor do they fully account for anticipated effects of multiple stressors. For example, an integrated watershed approach [@EPA96] has been utilized since the early 1990s to diagnose and manage water quantity and quality problems by addressing issues within hydrologically-defined geographic areas. Additionally, the habitat mosaic approach [@Henningsen05] of including multiple habitat types within restoration projects has been recognized as an effective means of allowing ecosystem state changes in response to different environmental pressures [@Duarte09; @Palmer09]. Adaptive management [@Holling78; @Gregory06] components have also been used to address challenges of sea-level rise, climate change, and development stressors, including monitoring to identify critical restoration decision points and needed intervention with contingency plans. Elements of each of these approaches could be combined to create a more holistic approach to guide restoration and conservation activities for coastal habitats in urban settings.

The Tampa Bay watershed (Florida, USA) is a valuable case study for developing a habitat restoration plan that addresses pervasive coastal stressors. Compared to other estuaries, the ratio of watershed to estuary area is small and the area is heavily developed with 42% of land use classified as urban and suburban residential [@SWFWMD18]. A retrospective approach to setting habitat protection and restoration targets in Tampa Bay was previously used [@Lewis96; @Robison10; @Cicchetti11; @Russell15]. Priority was given to restoration activities focused on habitat types that were important for a suite of estuarine faunal guilds disproportionately lost or degraded compared to a benchmark period. Criticisms of this approach included lack of consideration for future sea-level rise and other climate change factors [@Yoskowitz15], use of expanded and different habitats outside the Tampa Bay watershed [@Robison10], lack of attention to upland or freshwater wetland habitats, and little recognition of land development trends or actual available space for restoration efforts. These challenges are shared by restoration practitioners in other coastal environments and an approach that accommodates these challenges for planning would be highly transferable.

This paper describes an approach for habitat restoration and conservation planning that addresses the above challenges by considering the whole watershed, addressing historical changes, focusing on trajectories that have occurred during contemporary time periods, and considering both current and future stressors -- particularly land development and sea-level rise. Current and historical data are available for most Tampa Bay habitats, representing a time period when federal, state and local regulations were in effect and regional impacts from climate change have been documented [@Raabe12; @Cavanaugh14]. The approach establishes a framework that can guide both watershed-level habitat planning and site-level restoration activities and incorporates applicable elements of other habitat restoration paradigms discussed above [@Palmer09]. The general approach includes 1) designation of habitat types by strata relative to the aquatic-terrestrial gradient, 2) quantification of historical trends by habitat types to identify appropriate future targets in coverage, and 3) identification of opportunity areas that could be used by practitioners to achieve restoration goals based on habitat type and past trajectories. The outcomes are spatially-specific by providing maps to identity opportunity areas and reproducible using an open science workflow [@Lowndes17] that allows regular updates as new data become available.

@@

Tampa Bay is a large open water estuary (surface area approximately 983 km^2^) on the west-central coast of Florida (@fig-lulcmap). The watershed covers approximately 5,872 km^2^, for a total combined area of approximately 6,855 km^2^. The climate is subtropical and within the 2020 ecotone for mangrove and salt marsh habitats. Naive habitats in the watershed include pine flatwoods, forested freshwater wetlands and non-forested vegetated wetlands. The watershed is heavily developed with an estimated population of 3.3 million people in the four major counties [@Rayer20]*.* Numerous anthropogenic changes have altered the natural habits of Tampa Bay, including direct removal of habitat (e.g., dredge and fill of bay bottom, mining activities), alteration of hydrology, and destruction and fragmentation of habitat from development.

@@

The major habitat types of Tampa Bay were stratified by tidal influence and location in the watershed to define broad categories for restoration planning. Subtidal habitats included those that are submerged all or most of the time, intertidal habitats included emergent tidal wetlands that are submerged during high tides but exposed during low tides, and supratidal habitats included those that occur above the high tide line (i.e., the remainder of the watershed).

Subtidal habitats included hard bottom [@Jaap90; @Ash05], artificial reefs [@Dupont08], tidal flats [@Eisma98; @Moore68]*,* seagrasses [@Heck03; @Sherwood17], and oyster reefs [@Coen07; @zuErmgassen13]. Intertidal habitats (or emergent tidal wetlands) included mangroves [@Odum90], salt marshes [@Raabe12; @Comeaux12], salt barrens [@Berntness85; @Hsieh04]*,* tidal tributaries [@Sherwood08; @Wessel22], and living shorelines [@Smith18; @NOAA15; @RAE15]. Supratidal habitats included non-developed uplands [@Meyers90], freshwater forested wetlands [@Conner07], and freshwater non-forested wetlands [@Kushlan90]*.* Uplands were further sub-divided into coastal and non-coastal uplands based on location relative to the 5-foot contour that extended landward from the bay shoreline (described below). This 5-foot contour is an area of intense urban development and is expected to be impacted by sea-level rise based on current estimates. Since 1946, the St. Petersburg tidal gauge (NOAA gauge 872650) has documented a nearly 20 cm increase in mean tidal height to present day. Projections to the year 2100 suggest sea levels can increase between 58 and 259 cm in the region [@Burke19].

@@

Coverage targets for habitat types and opportunity areas for restoration were identified by integrating multiple datasets available for the region. First, habitat status and historical trends were quantified using land use/land cover and subtidal datasets to understand relative changes that have occurred over time. Second, historic habitat restoration efforts conducted in the watershed were synthesized as a guide for understanding a feasible level of effort for future projects that could be conducted by restoration practitioners. The first two steps were used to identify short-term (2030) targets and long-term (2050) goals for coverage (hectares) of native habitats. Finally, remaining opportunities for restoration were spatially identified by combining current coverages with existing or proposed protected areas and areas anticipated to be impacted by sea-level rise. As such, the approach identifies goals and targets and provides spatially explicit information that identifies where restoration practitioners should pursue projects to meet these goals and targets.

@@

For the majority of subtidal, intertidal and supratidal habitats, coverages were quantified from two routine spatial assessment programs conducted by the Southwest Florida Water Management District (SWFWMD). For subtidal habitats, 2018 data were used to estimate current coverage of seagrasses, tidal flats, and oysters [@SWFWMD19]. Historical datasets for subtidal habitats began in 1988 with updates occurring on an approximate bi-annual basis [@Sherwood17; @Tomasko20]*.* Oyster bed coverage has been routinely estimated in these data products beginning in 2014.

Current intertidal and supratidal habitat coverages were estimated using the 2017 SWFWMD Land Use Land Cover (LULC) maps [@SWFWMD18]. Land use and cover types (natural and developed) are classified following the Florida Land Use Cover and Forms Classification System [FLUCCS, @FDOT99; @SWFWMD14]*.* Mangroves, salt barrens, and salt marshes were reported individually. While the photointerpretation of specific freshwater wetland types is often difficult, forested wetlands and non-forested wetlands can be distinguished with these data. Therefore, all applicable FLUCCS codes representing natural freshwater wetlands were combined for these classifications. Native upland habitats were also combined in one classification. Historical estimates for all intertidal and supratidal habitats were also quantified starting with the earliest database in 1990 and occurring every two to three years until the current estimate in 2017.

To address data gaps for habitats not included in the routine SWFWMD datasets, results from special studies were compiled to obtain current estimates. These included hard bottom subtidal habitats, artificial reefs, tidal creeks, and living shorelines.  No information on historical trends is available for these habitats.

Finally, a habitat coverage change analysis between the terminal years of data (1988 to 2018 for subtidal, 1990 to 2017 for intertidal and supratidal) was conducted to understand how habitats were changing between types. This required an intersection of the data layers to quantify if habitat types were unchanged or changed for any given location and identifying the type of change (e.g., seagrass to tidal flats). The results were summarized as Alluvial diagrams showing relative proportions of habitat change by type and between years [@Allaire17].

@@

Restoration projects conducted over the past 40 years were quantified for each of the major habitat types to inform expectations for setting short-term targets and long-term goals. Restoration was defined as any activity that involved earthwork to reshape the land or the addition of structural elements (e.g., rock). Enhancement was defined as any activity not including earthwork that improved the environment (e.g., planting native vegetation, invasive species or debris removal, prescribed burns, etc.). Data were gathered from the SWFWMD Surface Water Improvement and Management Program, Federal Government Performance and Results Act reporting, the Tampa Bay Water Atlas (<https://www.tampabay.wateratlas.usf.edu/>), Tampa Bay Watch, and the Technical Advisory Committee of the Tampa Bay Estuary Program. The collected data included project name, year, description, size (area or length), and location (latitude and longitude). Data gaps were supplemented by archival research, site visits, contacting entities, and expert knowledge from local professionals.

@@

Spatially-explicit estimates of the opportunity areas and their restoration potential in the Tampa Bay watershed were obtained using a spatial analysis shown in @fig-gisworkflow. The two main processes included 1) binning existing datasets into relevant categories and 2) overlaying multiple datasets to identify opportunities. Opportunity areas were defined as locations where habitat protection and restoration activities are possible and where they could occur to attain the targets and goals described above. Identifying opportunity areas is necessary to quantify the restoration potential for a particular habitat type, which is a measure of what is actually possible given expected land use change and sea-level rise.

The Land Use Land Cover 2017 dataset from SWFWMD was used for binning existing coverages into the relevant habitat types in the intertidal and subtidal strata. All FLUCCS classification codes were placed into one of three categories. First, native habitats were those that included the full range of natural plant communities and other habitats that are endemic to the watershed Second, restorable habitats included existing altered but non-hardened and pervious FLUCCS codes that could potentially support native habitats through restoration. Third, existing development included developed land FLUCCS codes that are hardened and impervious (e.g., structures and pavement) and not suitable for habitat restoration.

After binning, the native and restorable lands were overlaid with additional layers to identify 1) reservation native and reservation restorable areas, and 2) existing and proposed native and restorable areas. Collectively, each of these unique products are considered the opportunity areas in the Tampa Bay watershed. The reservation native and reservation restorable areas are native and restorable habitats, respectively, that occur in the 5-foot contour or coastal stratum. This area extends from the local Mean Lower Low Water (MLLW) elevation to an elevation 5 feet above Mean Sea Level and is highly vulnerable to sea-level rise [@Burke19]. Native habitats in this stratum were identified as those to be reserved, whereas restorable habitats were identified as those where tidal wetlands could be restored.

The existing and proposed native and restorable areas were those that occurred in public lands that are currently acquired or proposed for acquisition. To identify these areas, native and restorable lands were intersected with data created from the Florida Natural Areas Inventory and inventorying conservation and drainage easements data. This workflow created the existing conservation and proposed conservation layers in @fig-gisworkflow . Intersecting the native and restorable lands in these areas produced four unique opportunity areas: existing conservation native, proposed conservation native, existing conservation restorable, and proposed conservation native.

@@

Compared to vegetation communities, soil characteristics typically change slowly (e.g., decades to centuries) in response to hydrologic impacts, unless physically disturbed [@Osland12; @Stockmann14]. Therefore, soil distributions can be used to estimate historical habitat distribution and restoration potential. A soils suitability layer was used for the Tampa Bay watershed [@Ries14] that classified soils as xeric, mesic, or hydric. The mesic and hydric categories were combined to represent wetland restoration potential and the xeric category was used to represent upland restoration potential. A distinction was made between tidal and freshwater wetland restoration potential by intersecting the mesic and hydric soils with the coastal stratum. Mesic or hydric soils that occur below the 5-foot contour were assigned a restoration potential for tidal wetlands, whereas mesic or hydric soils above the 5-foot contour were assigned a restoration potential for freshwater wetlands.

Two distinct mapping products were created from the above analysis. The first was an opportunities map that showed areas in the watershed identified as existing conservation native, existing conservation restorable, proposed conservation native, proposed conservation restorable, reservation native, and reservation restorable. The second was a map that identified the restorable lands (either existing or reservation) based on their restoration potential as coastal uplands, freshwater wetlands, native uplands, or tidal wetlands. All analyses were conducted using the R statistical programming language [@RCT22], specifically leveraging functions from the tidyverse package for data wrangling [@Wickham19] and the simple features (sf) package for geospatial analysis [@Pebesma18]. The workflows and data are provided in an open-access repository available on GitHub (<https://github.com/tbep-tech/hmpu-workflow>) [@Beck22].
